Investigating Fire Incidents Severity Levels in Toronto: Enhancing Resource Allocation for Fire Response

### 2024-12-11 paper/paper.qmd
Adjust reference to metholodgy appendix

## **Target Population**

@@

In the part of predicting fire severity, this study’s findings will be based on observational data from real fire incidents that occurred in Toronto. Therefore, there will not necessarily be sampling as in traditional surveys to collect data. Data will be gathered from official fire department records over a time period of 10 years. This time frame is chosen to ensure both inclusivity and relevance to the current state. The dataset will be divided into strata, subgroups of the population that have similar characteristics \[\@bailey1991strata\], based on different building types, with each stratum containing corresponding information, including fire origins, operational states of fire infrastructure systems, severity-related variables, and actions taken by the department.

A survey will be used to understand the current fire infrastructure system. I will employ the stratified sampling method. Stratified sampling involves dividing the target population into distinct strata based on shared characteristics \[\@neyman1992two\], in this case, it would be building type, as explained in the methodology overview paragraph. From each subgroup, a representative sample will be randomly selected to ensure that all relevant perspectives are included. All citizens involved in fire incidents (for example, people caught in the fire, building managers, and neighbours) will have a chance to fill out a form. The survey will inquire about the presence, timeliness, effectiveness, and ease of use of fire infrastructure and fire-reducing facilities, such as public fire extinguishers and fire water taps.

@@

### **Fire Incident Records:**

@@

<!-- -->

@@

-   **Privacy Protection**: Privacy of participant information and data is important to all observational data collection \[\@preibusch2013guide\]. To protect the privacy of those involved, the data will be anonymized. Specific addresses will be removed to prevent the identification of individuals or locations. Additionally, rare fire incident data will only be included after obtaining the consent of the individuals involved, ensuring their participation is voluntary.

### **Survey on Real Experience with Fire Infrastructure Systems:**

-   Surveys will be distributed using a stratified sampling method to individuals involved in fire incidents, ensuring coverage across all levels of severity and building types. These surveys will gather information on the effectiveness of fire response systems, such as fire alarms, sprinkler systems, and smoke detection systems, in reducing damage and protecting properties. The responses will help guide future improvements in fire infrastructure.

-   **Participant Recruitment:**

    -   Individuals who were directly involved in or reported the fire incident will be asked to provide their contact information, such as phone numbers, to ensure can reach them for participation. In general, building managers will distribute the survey to all individuals within the building who may have experienced or been impacted by the fire incident. This ensures the study reaches a wide range of individuals with relevant insights.

    -   Since fire incidents involve people’s safety, participants are likely to be motivated to share their experiences and provide valuable feedback on fire operations and infrastructure. As a way to express appreciation, participants who complete the survey will be given the opportunity to enter a raffle for a prize, which will be drawn at the end of each year. This incentive is designed to encourage engagement while ensuring that participation remains voluntary \[\@freeman2007standards\].

    -   All Participants will also be directed to a Toronto Fire Prevention website (<https://www.toronto.ca/community-people/public-safety-alerts/fire-prevention-inspection-enforcement/>) [@toronto_fire_prevention_2024], where they can access additional information and resources related to fire safety.

-   **Survey Design:**

    -   The survey is designed to gather both qualitative and quantitative data regarding fire safety infrastructure and its effectiveness. The questions are phrased in a neutral, unbiased manner, and all multiple-choice questions will include a "does not apply" or “other” option to ensure inclusivity and clarity for all respondents. To minimize the risk of trauma triggerings \[\@alessi2023toward\], participants will be informed at the beginning of the survey that they may choose not to answer any questions that make them uncomfortable. A reminder of available support resources will be provided at the end of the survey. Participants will also be reminded of their right to quit the survey at any time without any consequences.

@@

## **Data Validation**

@@

**Missing Data**: If a record is missing important information, such as the area of the fire, building details, or TFS response personnel, the record will be removed to ensure the quality of the dataset. In cases where operational system information is missing, for example, if the TFS report doesn’t specify details on the operation of fire systems of an incident, I will cross-check with survey responses to find the missing data for that specific incident. If there is inconsistency in the fire system status, such as conflicting responses between the survey and TFS data, or if no survey response is available for a particular incident, the operational data will be classified as "unreported" \[\@stavseth2019handling\].

**Category Validation**: To ensure all recorded information aligns with clearly defined and consistent categories. This includes verifying that the fire incident categories, including severity levels, fire origin, and response actions are accurately defined and organized according to standard protocols. I will review the definitions and ensure that terminology is used consistently across all data sources. Additionally, I will ensure that the severity of the fire is defined in a way that is consistent with fire severity classification standards specific to Toronto. For example, the Office of the Fire Marshal of Ontario has set standards for fire prevention and reporting \[\@ontario_fire_marshal\].

## **Modeling - Severity Prediction and Recommendations for Agencies and the Public**

### **Model - Random Forests**

-   The Random Forests model will be used to identify key predictors of fire severity. These models are an ensemble of multiple decision trees, which together provide predictions by combining the results of various decision paths for critical variables under certain thresholds \[\@breiman2001random\]. The model will identify important factors influencing fire severity, and the predictions will include the weighting of factors based on the frequency and severity of past incidents. For example: even though high-severity fires may be less frequent, understanding the factors leading to these incidents is crucial, so weight adjustments will ensure that less frequent but high-impact factors are appropriately considered in the predictions.

### **Model validation**

-   To ensure the models generalize well and prevent overfitting, cross-validation techniques will be applied, this the total data will be separated into subsets, and the model will be rained on some subsets, and tested on others \[\@berrar2019cross\].  This will increase the reliability of the model. The model performance will be assessed using evaluation metrics such as ROC curves \[\@hoo2017roc\] which is a visual demonstration of the model’s sensitivity an specificity level. Other evaluation criteria include accuracy scores will also be included to ensure that the model’s predictions are reliable and effective.

### **Applications and Benefits**:

-   The results from the Random Forest model will inform fire agencies in predicting potential fire incidents. By analyzing the time of year and other features, the model can help agencies prepare for periods when high-severity fires are more likely. Additionally, the model can provide guidance on resource allocation. For example, if in a fire incident call, people indicate that a fire originated in a factory near electrical machinery, and the model predicts a high-risk fire in that area, the fire department can allocate more resources to manage the situation effectively. On the other hand, if a fire occurs in a residential building with the origin in a bathroom, and the model predicts a lower severity, fewer resources can be allocated, leading to more efficient use of time and resources.

-   Knowledge gained from the model can be used to educate the public about areas with higher risks of severe fires. By targeting fire prevention measures and raising awareness in these high-risk locations, the city’s overall fire safety can be improved, reducing damage and loss of life. For example, if fire origins like HVAC systems or electrical equipment, are discovered to be strongly linked to high-severity fires, the fire agency can prioritize fire prevention plans in these areas.

## **Survey Analysis - Effectiveness of Fire Infrastructure and Future Improvements**

The survey can gather feedback on the presence, timeliness, effectiveness, and ease of use of various fire infrastructure and fire-reducing facilities, such as fire extinguishers, fire hoses, sprinkler systems, and smoke detectors \[\@ \[\@thomas2002effectiveness\]\]. The responses will help identify areas where improvements can be made and assist agencies in understanding the current state of fire safety infrastructure.

-   A combination of survey responses and records maintained by Toronto Fire Services (TFS) will provide valuable insights into the effectiveness of existing fire systems. By analyzing the data through categorizing responses; identifying geographical patterns; correlating feedback with fire severity; and analyzing open-ended responses, I can identify which systems are working well and where additional resources or improvements may be needed. This could include ensuring buildings are equipped with adequate alarm systems, performing regular maintenance on sprinklers, and enhancing other fire safety features.

### **Applications and Benefits**:

@@

### Link to the Survey:

Here is a link to constructed survey: [https://forms.gle/gQcRL3YTL52S9xgR6](#0)

>>>>

## Target Population

@@

In the part of predicting fire severity, this study’s findings will be based on observational data from real fire incidents that occurred in Toronto. Therefore, there will not necessarily be sampling as in traditional surveys to collect data. Data will be gathered from official fire department records over a time period of 10 years. This time frame is chosen to ensure both inclusivity and relevance to the current state. The dataset will be divided into strata, subgroups of the population that have similar characteristics [@bailey1991strata], based on different building types, with each stratum containing corresponding information, including fire origins, operational states of fire infrastructure systems, severity-related variables, and actions taken by the department.

A survey will be used to understand the current fire infrastructure system. I will employ the stratified sampling method. Stratified sampling involves dividing the target population into distinct strata based on shared characteristics [@neyman1992two], in this case, it would be building type, as explained in the methodology overview paragraph. From each subgroup, a representative sample will be randomly selected to ensure that all relevant perspectives are included. All citizens involved in fire incidents (for example, people caught in the fire, building managers, and neighbours) will have a chance to fill out a form. The survey will inquire about the presence, timeliness, effectiveness, and ease of use of fire infrastructure and fire-reducing facilities, such as public fire extinguishers and fire water taps.

@@

### Fire Incident Records:

@@

@@

-   **Privacy Protection**: Privacy of participant information and data is important to all observational data collection [@preibusch2013guide]. To protect the privacy of those involved, the data will be anonymized. Specific addresses will be removed to prevent the identification of individuals or locations. Additionally, rare fire incident data will only be included after obtaining the consent of the individuals involved, ensuring their participation is voluntary.

### Survey on Real Experience with Fire Infrastructure Systems:

Surveys will be distributed using a stratified sampling method to individuals involved in fire incidents, ensuring coverage across all levels of severity and building types. These surveys will gather information on the effectiveness of fire response systems, such as fire alarms, sprinkler systems, and smoke detection systems, in reducing damage and protecting properties. The responses will help guide future improvements in fire infrastructure.

#### Participant Recruitment:

Individuals who were directly involved in or reported the fire incident will be asked to provide their contact information, such as phone numbers, to ensure can reach them for participation. In general, building managers will distribute the survey to all individuals within the building who may have experienced or been impacted by the fire incident. This ensures the study reaches a wide range of individuals with relevant insights.

Since fire incidents involve people’s safety, participants are likely to be motivated to share their experiences and provide valuable feedback on fire operations and infrastructure. As a way to express appreciation, participants who complete the survey will be given the opportunity to enter a raffle for a prize, which will be drawn at the end of each year. This incentive is designed to encourage engagement while ensuring that participation remains voluntary [@freeman2007standards].

All participants will also be directed to a Toronto Fire Prevention website (<https://www.toronto.ca/community-people/public-safety-alerts/fire-prevention-inspection-enforcement/>), where they can access additional information and resources related to fire safety [@toronto_fire_prevention_2024].

#### Survey Design:

The survey is designed to gather both qualitative and quantitative data regarding fire safety infrastructure and its effectiveness. The questions are phrased in a neutral, unbiased manner, and all multiple-choice questions will include a "does not apply" or “other” option to ensure inclusivity and clarity for all respondents. To minimize the risk of trauma triggerings [@alessi2023toward], participants will be informed at the beginning of the survey that they may choose not to answer any questions that make them uncomfortable. A reminder of available support resources will be provided at the end of the survey. Participants will also be reminded of their right to quit the survey at any time without any consequences.

@@

## Data Validation

@@

**Missing Data**: If a record is missing important information, such as the area of the fire, building details, or TFS response personnel, the record will be removed to ensure the quality of the dataset. In cases where operational system information is missing, for example, if the TFS report doesn’t specify details on the operation of fire systems of an incident, I will cross-check with survey responses to find the missing data for that specific incident. If there is inconsistency in the fire system status, such as conflicting responses between the survey and TFS data, or if no survey response is available for a particular incident, the operational data will be classified as "unreported" [@stavseth2019handling].

**Category Validation**: To ensure all recorded information aligns with clearly defined and consistent categories. This includes verifying that the fire incident categories, including severity levels, fire origin, and response actions are accurately defined and organized according to standard protocols. I will review the definitions and ensure that terminology is used consistently across all data sources. Additionally, I will ensure that the severity of the fire is defined in a way that is consistent with fire severity classification standards specific to Toronto. For example, the Office of the Fire Marshal of Ontario has set standards for fire prevention and reporting [@ontario_fire_marshal].

## Modeling - Severity Prediction and Recommendations for Agencies and the Public

### Random Forests

The Random Forests model will be used to identify key predictors of fire severity. These models are an ensemble of multiple decision trees, which together provide predictions by combining the results of various decision paths for critical variables under certain thresholds [@breiman2001random]. The model will identify important factors influencing fire severity, and the predictions will include the weighting of factors based on the frequency and severity of past incidents. For example: even though high-severity fires may be less frequent, understanding the factors leading to these incidents is crucial, so weight adjustments will ensure that less frequent but high-impact factors are appropriately considered in the predictions.

### Model Validation

To ensure the models generalize well and prevent overfitting, cross-validation techniques will be applied, this the total data will be separated into subsets, and the model will be rained on some subsets, and tested on others [@berrar2019cross]. This will increase the reliability of the model. The model performance will be assessed using evaluation metrics such as ROC curves [@hoo2017roc] which is a visual demonstration of the model’s sensitivity an specificity level. Other evaluation criteria include accuracy scores will also be included to ensure that the model’s predictions are reliable and effective.

### Applications and Benefits:

The results from the Random Forest model will inform fire agencies in predicting potential fire incidents. By analyzing the time of year and other features, the model can help agencies prepare for periods when high-severity fires are more likely. Additionally, the model can provide guidance on resource allocation. For example, if in a fire incident call, people indicate that a fire originated in a factory near electrical machinery, and the model predicts a high-risk fire in that area, the fire department can allocate more resources to manage the situation effectively. On the other hand, if a fire occurs in a residential building with the origin in a bathroom, and the model predicts a lower severity, fewer resources can be allocated, leading to more efficient use of time and resources.

Knowledge gained from the model can be used to educate the public about areas with higher risks of severe fires. By targeting fire prevention measures and raising awareness in these high-risk locations, the city’s overall fire safety can be improved, reducing damage and loss of life. For example, if fire origins like HVAC systems or electrical equipment, are discovered to be strongly linked to high-severity fires, the fire agency can prioritize fire prevention plans in these areas.

## Survey Analysis - Effectiveness of Fire Infrastructure and Future Improvements

The survey can gather feedback on the presence, timeliness, effectiveness, and ease of use of various fire infrastructure and fire-reducing facilities, such as fire extinguishers, fire hoses, sprinkler systems, and smoke detectors [@thomas2002effectiveness]. The responses will help identify areas where improvements can be made and assist agencies in understanding the current state of fire safety infrastructure.

A combination of survey responses and records maintained by Toronto Fire Services (TFS) will provide valuable insights into the effectiveness of existing fire systems. By analyzing the data through categorizing responses; identifying geographical patterns; correlating feedback with fire severity; and analyzing open-ended responses, I can identify which systems are working well and where additional resources or improvements may be needed. This could include ensuring buildings are equipped with adequate alarm systems, performing regular maintenance on sprinklers, and enhancing other fire safety features.

### Applications and Benefits:

@@

## Link to the Survey:

Here is a link to constructed survey: <https://forms.gle/gQcRL3YTL52S9xgR6/>
